---
title: "Interactive Plot using plotly for GDP Vs Life Expectancy"
author: "Nikitha Agarampalli"
date: "April 18, 2018"
output: html_document
---

```{r global_options, include=FALSE}
rm(list=ls()) 
library(knitr)
opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
               echo=TRUE, warning=FALSE, message=FALSE)
```

## Overview  
Interactive data visualization has been a pursuit of any data analytics person or statistician as it enables direct action on a plot to perform any changes and derive conclusions based on it. This type of plots allow deeper exploration of data than static plots. 

The goal of this assignment is to create an interactive plot based version of the nations.csv dataset to show GDP Vs Expectancy results.  

```{r}
# Loading the nesessary packages

library(plotly)
library(readr)

```

## Reading the data and plotting it  

```{r}

# Reading the data file
nations_data <- read_csv("nations.csv", col_types = cols(population=col_double(),  year = col_date(format="%Y")))

# Foreground data
nations_data_targets<- subset(nations_data, country=="Ireland" | country =="Norway" | country =="Sweden"   )

# Background data = complement set 
nations_minus_targets<- subset(nations_data, country!="Ireland" & country !="Norway" & country !="Sweden"  )

# Your background layer
p_nations<- ggplot(nations_minus_targets, aes(x =year, y=gdp_percap, group = country)) + 
# this theme clears away grid lines, makes backgound white  
theme(panel.grid.major = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black", size = 0.25), axis.title.x=element_blank(), axis.text.x = element_text(angle = 30, hjust=1, vjust = .5), legend.key = element_rect(fill = NA, colour = NA, size = 0.25)) 

p_nations<- p_nations + geom_line( size= 0.25, na.rm = TRUE, color="grey90", alpha =0.6, 
                                   show.legend = FALSE ) + geom_line(data=nations_data_targets, 
                                                                    size =1, show.legend = TRUE,                                                                            (aes(x=year, y=gdp_percap,                                                                              colour=country, group=country))) +

                        scale_colour_brewer(palette = "Set1") + 
                        ggtitle("1990-2014 : GDP per capita for Ireland, Norway & Sweden")
p_nations

```

## Interactive plot of the above graph  
The above graph is plotted here using plotly with a range slider.  

```{r}

g <- nations_data %>% 
  
  # Group by country
  # initiate a plotly object with date on x and median on y & make the lines grey
  
  group_by(country) %>% plot_ly(x = ~year, y = ~gdp_percap) %>%
  add_lines(name = "Other Nations", hoverinfo = "none", type = "scatter", mode = "lines", 
            line = list(color = 'rgba(192,192,192,0.4)')) %>%
  
  # Plot separate lines for Dallas & Houston. Give the lines different colours and make them hoverable
  
  add_lines(name = "", data = filter(nations_data, country %in% c("Ireland", "Norway", "Sweden")),
            hoverinfo = "country", line = list(color = c("red", "blue")), color = ~country) %>%
  layout(xaxis = list(rangeselector = list(buttons = list(list(count = 3, step = "year",
                                                               steepmode = "backward"),
                                                          list(count = 6, step = "year",
                                                               stepmode = "backward"),
                                                          list(count = 1, step = "year", 
                                                                stepmode = "backward"), 
                                                          list(count = 1, step = "year",
                                                               stepmode = "todate"),
                                                          list(step = "all"))),
                 rangeslider = list(type = "Year")), yaxis = list(title = "GDP Per Capita"))  
g

```